Applying gamification to developer practices

### 2013-09-04 paper_id1.tex
progress is slow but steady

\title{Effectiveness of Instant Feedback on Developer Navigation Practices}
\subtitle{Does Instant Feedback Make Us Better Developers?}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

Based on the survey results we draw the following conclusions:
•	95\% of developer responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current workflow pattern the developer follows.
•	Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.  So, the tool will incorporate team competition ideas and recognize quality practices.
•	Over 90\% of respondents approved of sharing data with their team members

\section{Study Design}
Creating a longitudinal study involves defining a long-term goal that is measurable in increments, determining the discrete steps that the study will use to communicate the goal to the participants, distribute interventions to participants, and track participants progress over time.  Many software engineering studies are conducted at discrete points in time via surveys of sampled participants.  Perhaps prior to an intervention, and multiple points following an intervention.  This study demonstrates a method to perform an intervention, continuously monitor the progress of participant response to that intervention, and provide continuous feedback to the participant and researchers about the progress.  Continuous meaning that no information gaps exist in the data collection, there are no periods of unmonitored activity by the participant.  Continuous also applies to the intervention being evaluated, where the participant and researchers receive constant feedback about the conformance to the intervention's purpose.  Studying an intervention in a continuous context determines how sticky the intervention is in the mind of the developer.  It can remove bias that might be introduced by periodic surveys or audits and enhance the habits formed in the participant receiving the intervention.

Designing the study with continuous measurement but no control group presents a traditional study design with some challenges.  We overcame this by dividing the study into phases where the participants themselves form a control group.  The first phase baselines their activities determining the current state of their practices.  The second phase presents the intervention to the participants.  This phase the participants do not receive continuous feedback on their compliance with the intervention.  This phase simulates the traditional approach of communicating an intervention then returning with an audit to determine whether participants are complaint with it.   The last phases turns on continuous feedback to the participants on their compliance with the intervention.  As they receive the feedback, the participants can determine for themselves how much they are implementing the intervention.  In the last phase they possess complete knowledge of the intervention and their compliance with it making it plausible that they can continuously control their compliance.

@@

Gamification motivates people to achieve goals through incremental scoring and discrete awards that are typically shared with peers in an on line community.  Companies are gamifying their business processes and even public facing web sites to encourage people to complete a task such as their profile, stay loyal by earning rewards (air miles), or participate frequently by being part of a community.  Aspects of Gamification that should be considered when applying this popular technique to a process or system include motivational alignment, whether to include the competitive dimension, and making achievements encouraging.

First it is important to consider the existing motivation people may have for performing the activity.  Applying gamification to activities that people are motivated intrinsically to do may decrease their motivation.  For example applying gamification to education is very difficult because learners are typically intrinsically motivated more than extrinsically motivated to achieve in their education \cite{wbsnipes:Vansteenkiste2008Does}.    In other words they are mostly interested in learning to better themselves rather than earn the highest GPA.   If an activity is primarily extrinsically motivated then gamification is an easy fit to directly apply to the activity.

Several studies support the use of games as a learning aid or as the core part of a training program.  Naicker, Vincent and Adams \cite{wbsnipes:Amory1999Use} studied which type of computer game had the best response for enjoyment of play among undergraduate students.  The authors concluded that adventure games provide the best foundation for development of teaching resources.  Students rated sound, graphics, story line, and use of technology as important game elements.

@@

Using these guidelines we assigned points to structured navigation commands which received 1-10 points each while unstructured navigation commands received 0 points.  Assigning negative points is discouraged as this may demotivate folks who do not have detailed knowledge of how the points are assigned.   In creating levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we built levels following an exponential curve.  At first users level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve  making the game motivating for beginners and challenging for more experienced developers.  We designed the points and levels such that above average developers can complete all levels during the study period.

@@

After selecting the practice to focus on, we needed to select an intervention method.  The term intervention, indicates that we are seeking to influence developers to perform a practice previously defined by software engineering community.  Intervention involves communication about the practice in a way that motivates adoption \cite{}.  Interventions can include presentations, one-on-one discussions, distance education mechanisms such as webinars, and word of mouth recommendations.  Part of the research objective is to determine whether we can achieve adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.  Utilizing the automation capabilities of Blaze, we chose to automatically pop-up a window with training information on structured navigation practices and tools. For this study we chose to make a nice looking graphic that users click on to see information about navigation commands, obtain and install the Sando search tool, and learn about the Blaze game.  The site uses a combination of video and blog-post style information easily accessible to the developer.  The usefulness of the communication site is rated as part of our study completion survey.

The study established three time periods, a preliminary period to establish a baseline, a period where information is communicated but no immediate feedback or points are displayed, and a period where points are displayed and all information is available.  We chose to deploy the Blaze tool without communicating the study purpose for the baseline period so that an accurate picture of the baseline could be captured by the tool.  Then the tool would pop-up a display with the information on the communication web site asking the developers to consume the information and begin using the practices.   Finally the display of points fo accumulated by the developer provides the instant feedback mechanism.

@@

Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository.\cite{Singer2012It}  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.  Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.  The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.  Results show an increase in the frequency of commits at a statistically significant level.  This study uses finer grained instrumentation to identify practices being used between commits such as structured navigation.

@@

Building on this success, this work proposes a generalized method to automate the identification of the techniques described in the related work and others, and determine when the techniques would help a developer improve their work patterns.  Then we intend to create a game-like virtual environment where developers learn from each other and are motivated to adopt the best practices developed by these and other researchers.

>>>>

\title{Applying gamification to developer practices}
\subtitle{Why developer habits die hard}
% 
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

Based on the survey results we made the following conclusions:
•	95\% of developer responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current workflow pattern the developer follows.
•	Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.  So, the tool will incorporate team competition ideas and recognize quality practices.
•	Over 90\% of respondents approved of sharing data with their team members

\section{Study Design}

Utilizing the unobtrusive continuous monitoring capabilities of Blaze, we designed the pilot as a longtiudinal study in three waves.  Longtiudinal stuides allow us to design an experiement and determine the effects over two or more sample periods using standard MANOVA (Multiple ANalysis of VAriance) models. \cite{longitudinalbook}  We employed the three wave model defining the first wave as a baseline capturing the current practices of the pilot group.  The second wave employed the intervention that included an informative web site on structured navigation commands in Visual Studio, information about the sando search tool, and information about how points would be awarded during the pilot.  The third wave provided the participants with continuous feedback in the from of points accumulated and an indicator arrow \ref{fig:blazegui} that points up when their navigation ratio is higher than their historical average.

@@

@@

Using these guidelines we assigned points to structured navigation commands which received 1-10 points each while unstructured navigation commands received 0 points.  Assigning negative points is discouraged as this may demotivate folks who do not have detailed knowledge of how the points are assigned.\cite{}   In creating levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we built levels following an exponential curve.  Users initially level up after a working day or two extending to achieving the next level after a week or even a month following an exponential curve.  The leveling up points make the game motivating for beginners and challenging for more experienced developers.  We designed the points and levels such that above average developers can complete all levels during the study period.

@@

After selecting the practice to focus on, we needed to select an intervention method.  The term intervention, indicates that we are applying a technique to influence a change in the measured behavior of developers.  Intervention involves communication about the practice in a way that motivates adoption \cite{}.  Interventions can include presentations, one-on-one discussions, distance education mechanisms such as webinars, and word of mouth recommendations.  Part of the research objective is to determine whether we can achieve adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.

In the initial wave, participants were informed of the purpose of the study what to expect from the Blaze tool.  The participants installed the tool in their Visual Studio environment.  The tool did not present any window for them to see and simply collected data in the background.

After a week's data collection, Blaze started to automatically pop-up a window containing a button for a web page with information.  The page included a link to download another tool for code search called Sando \ref{sando} and information about using that tool in a video.  The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.  The About Blaze page contained details of the point scoring system and background on Blaze's purpose.  The usefulness of the communication site is rated as part of our study completion survey.

@@

Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository.\cite{Singer2012It}  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.  Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.  The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.  Results show an increase in the frequency of commits at a statistically significant level.  This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.

@@

Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development. \cite{Dubois2013Understanding} Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.  They provide guidance on assessing gamification in the software development domain that includes analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.  Dbois and Tamburrelli demonstrate this guidance using a study of students working on a class project.  Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.  The subject group can see each others' rankings while the control group only sees their personal rankings.  Results though inconclusive indicate that competition increases the desired results.  In our study we apply both individual feedback and competitive based feedback (anonymized) to participants and seek to observe a change in their behavior after an intervention.

\section{Results}
The following hypotheses interest us in the current study.  
1. Developers using blaze use more structured navigtion when they receive instant feedback about their practices
2. Developers find the competitive aspect of blaze encourages them to pay more attention to navigation practices

To evaluate whether developers use more structured navigation when they receive instant feedback, we monitor the change in navigation ratio over the pilot study period.   Expecting that as developers learn about navigation commands and tools that count towards the score in the second week they will begin using the practices more.  Also expecting that noticing the leaderboard competition will encourage some developers to use the navigation practices more.  What we observed is shown in Figure \ref{fig:navigationaverage} that the structured navigation practice did not increase significantly from the beginning to the end of the study or between weeks when the additional information was revealed to them.

\begin{figure}\begin{mdframed}[linecolor=white]
	\includegraphics[width=3in]{navigationaverage.pdf}
	\caption{Average Navigation Ratio by Week by Developer}
	\label{fig:navigationaverage}
\end{mdframed}\end{figure}

\section{Threats to Validity}
There are several threats to validity for this result.  By making the purpose of our measurement and contest known although without details we may have triggered the participants to think about navigation commands they use before establishing a baseline.  We attempt to control for this by comparing the pilot participants' data to data from Blaze users who are knowledgeable Visual Studio developers but did not receive the interventions.  Another threat is the intact team represents people with different backgrounds and experience levels who may have established habits for navigating through source code.  We attempt to control for this by asking post survey questions of the participants about their familiarity with the code they work in and their experience levels.  The threat of non- generalizable of results remains particularly because the pilot was conducted within an intact team in a single industrial site in India. As the pre-study questionnaire shows, the developer's desirability for a gamification type of intervention vary by country.
